Supplementary material for "Hack Weeks as a model for Data Science Education and Collaboration"

### 2017-12-28 sections/participants.tex

\section*{Audience and Participant Selection}

Hack weeks differ from traditional conferences or summer schools in that knowledge transfer occurs across many levels of seniority, disciplinary boundaries, and novelty of the topics discussed.
In addition, a substantial amount of hack week content is generated during the event itself, requiring active participation from participants.
Therefore in order to maximize learning outcomes and collaborative exchanges, it is crucial that the participant selection process be carried out with considerable care.

@@

No matter the selection procedures used, we encourage organizers to critically examine their cohort selection, experiment with new approaches, and routinely evaluate the different steps in their procedure%.  For example, merit scores can be interrogated for shared biases, and one can plot the demographics both before and after the initial selection step to diagnose whether certain groups are preferentially selected or excluded
(see also the Supplementary materials for more details on how the previous hack weeks approached this problem).

%One way to maximize transparency in the selection process is to minimize human decision making steps that introduce biases, and to . 
%We work to achieve this by first assessing the merit of each candidate with respect to the overall goals of the hack week.
%We try to minimize bias in this step by blinding ourselves to a candidate's other attributes, including name and other personal information, and assess their candidacy based soley on questions asked specifically for this purpose.
%When doing this procedure for a large enough sample, it is unlikely that the resulting pool of acceptable candidates is smaller than the number of available spaces at the workshop.

>>>>

\section*{Audience and Participant Selection}
\begin{bf}
Hack weeks differ from traditional conferences or summer schools in that knowledge transfer occurs across many levels of seniority, disciplinary boundaries, and novelty of the topics discussed.
In addition, a substantial amount of hack week content is generated during the event itself, requiring active participation from participants.
Therefore in order to maximize learning outcomes and collaborative exchanges, it is crucial that the participant selection process be carried out with considerable care.

@@

No matter the selection procedures used, we encourage organizers to critically examine their cohort selection, experiment with new approaches, and routinely evaluate the different steps in their procedure%.  For example, merit scores can be interrogated for shared biases, and one can plot the demographics both before and after the initial selection step to diagnose whether certain groups are preferentially selected or excluded
(see also the Supplementary materials for more details on how the previous hack weeks approached this problem). 
\end{bf}
%One way to maximize transparency in the selection process is to minimize human decision making steps that introduce biases, and to . 
%We work to achieve this by first assessing the merit of each candidate with respect to the overall goals of the hack week.
%We try to minimize bias in this step by blinding ourselves to a candidate's other attributes, including name and other personal information, and assess their candidacy based soley on questions asked specifically for this purpose.
%When doing this procedure for a large enough sample, it is unlikely that the resulting pool of acceptable candidates is smaller than the number of available spaces at the workshop.

### 2017-12-28 sections/why.tex

\begin{itemize}
\item{\textit{Education and Training}: 
While some hack weeks are focused more on education than others (see Figure \ref{fig:hackspectrum}), there is often a skill-development component that entails extensive discussion on reproducible research and open science practices. Participants gain a strong foundation in open science practices from the diverse group setting and go on the become ambassadors for such practices in their respective fields. This type of lateral knowledge transfer is a core attribute of a hack week, and provides an opportunity to learn skills that are not described in papers and software implementations.}

@@

\item{\textit{Community Building}: Hack weeks provide a tremendous opportunity to catalyze community development through a shared interest in solving computational challenges with open source software. These events allow computationally minded researchers to break from the isolation of their academic departments, build connections and spark new collaborations.}

@@

It is worth noting that the reasons for participants to attend a hack week are as diverse as the reasons for running such an event. 
Beginner participants may attend primarily to learn a new technique and may not have an explicit plan for project based work, while others may attend to gain experience in mentoring, or to focusing on an existing project already in progress.

>>>>

\begin{itemize}
\item{\textit{Education and Training}:
While some hack weeks are focused more on education than others (see Figure \ref{fig:hackspectrum}), there is often a skill-development component that entails discussions about reproducible research. Participants gain a strong foundation in open science practices from the diverse group setting and go on the become ambassadors for such practices. This type of lateral knowledge transfer is a core attribute of a hack week, and provides an opportunity to learn skills that are not described in papers and software implementations.}

@@

\item{\textit{Community Building}: Hack weeks provide a tremendous opportunity to catalyze community development through a shared interest in solving computational challenges with open source software. These events allow computationally minded researchers to break from the isolation of their academic departments and spark new collaborations.}

@@

It is worth noting that the reasons for participants to attend a hack week are as diverse as the reasons for running such an event.
Beginner participants may attend primarily to learn a new technique, while others may attend to gain experience in mentoring, or to focusing on an existing project already in progress (for more details on setting objectives, see supplementary materials).

### 2017-12-28 supplementary_material.tex

\subsection{Hack Week Location}
When considering space, both the broader location (city and country) as well as the specifics (the room in which the workshop will occur) should receive equal consideration. The non-traditional nature of hack weeks means that some students report challenges in receiving support from their supervisor or department to attend. Thus, keeping the costs of attendance low (including travel costs, room and board, and conference fees) should be a high priority when deciding on a location. Conference centers like the Lorentz Center at the University of Leiden provide funding along with the collaborative space, but have long lead times when applying for their spaces (typically 12-18 months). The same can be true for collaborative spaces within universities and active learning environments, which tend to be popular. Therefore, the search for a location should commence no later than twelve months before the workshop, ideally much earlier.
The space itself should be large and ideally configurable, with moveable tables and chairs, several screens or projectors, and an ample number of white-/chalkboards. A useful guiding principle is to book a space that fits a group at least 25\% larger than that envisioned for the workshop, and the less configurable the space, the larger the extra room should be.
There should be plenty of power outlets for participants to plug in their computers.
The space should have ready access to wifi, for activities that require an internet connection (e.g., collaborative software development), and organizers should arrange in advance to provide access to visitors from other institutions.
Accessibility should be taken into consideration: is the space accessible for participants who use wheel-chairs? Are accessible restrooms available.

@@

Event size (i.e. the number of attendees) and event durations are important to consider during the design and planning phase of a hack week. All hack weeks held so far have been five days, with group sizes between 40-60 participants. The number of attendees resulted from a number of different design considerations. One practical consideration was the size of the available location: especially the project-focused parts of a hack week generally require a reconfigurable space with more seats than attendees, such that spontaneous organization in sub-groups is easy. However, group size is also important in the context of community building, group dynamics and programme decisions. We limited group size to a maximum of 70 participants, because previous anecdotal evidence has shown us that this is about the maximum group size where interactions between all participants is still possible. We wanted to ensure that all participants had an opportunity to at least meet all other participants, in order to facilitate community building and keep the group from fracturing into cliques. Another important consideration is representation along the many axes of diversity we considered. For example, a single undergraduate participant might feel like an outsider; admitting three or four can mitigate alienation effects. Similarly, in order to foster active collaboration among many different subfields, we ensured admitting participants with many different types of expertise, some of which overlapped. Having a set of participants that is too small would interfere both with these goals of representation and of fostering interdisciplinary collaborations and networking, thus we would advise admitting at least $\sim$30 or so participants if these objectives are an important part of the hack week.

@@

Feedback is a crucial component of a hack week. Because of its experimental design, and the differences in group from year to year, it is important to routinely check whether the workshop still matches up with what participants expect and find useful. Additionally, computational fields tend to be fast-paced; a tutorial given one year might be outdated a few years later. It is thus useful to put some detailed thoughts into the survey. Following from discussions the committee has likely had during the initial stage of preparing the announcement and selecting participants, organizers will want to confront their own expectations before the workshop, and design a survey that allows for both measurements of specific outcomes and goals the organizers are interested in, as well as leave enough free space for participants to self-report outcomes that arose spontaneously or were out of scope of the initial design. It is helpful to design the survey early on and allow for some time near the end of the workshop for participants to fill it out there; in our experience, this leads to much higher response rates than if the survey is distributed after the end of the workshop. While some improvements and changes in the survey design from year to year are inevitable, it is advisable to set a number of core questions that do *not* change from year to year. These questions will allow organizers to assess the workshop's function within their community, and assess how changes to the format between one year and another may affect outcomes of interest.

>>>>

\subsection{Hack Week Location}
When considering space, both the broader location (city and country) as well as the specifics (the room in which the workshop will occur) should receive equal consideration. The non-traditional nature of hack weeks means that some students report challenges in receiving support from their supervisor or department to attend. Thus, keeping the costs of attendance low (including travel costs, room and board, and conference fees) should be a high priority when deciding on a location. Conference centers like the Lorentz Center at the University of Leiden provide funding along with the collaborative space, but have long lead times when applying for their spaces (typically 12-18 months). The same can be true for collaborative spaces within universities and active learning environments, which tend to be popular. Therefore, the search for a location should commence no later than twelve months before the workshop, ideally much earlier.
The space itself should be large and ideally configurable, with moveable tables and chairs, several screens or projectors, and an ample number of white-/chalkboards. A useful guiding principle is to book a space that fits a group at least 25\% larger than that envisioned for the workshop, and the less configurable the space, the larger the extra room should be.
There is a natural tension between keeping the group together and providing physically separated break-out spaces. On the one hand, conducting the workshop in a single large venue improves group cohesion and prevents self-segregation of researchers and ideas. On the other hand, the interactive nature of the workshop may lead to an environment where it is difficult for individuals to focus on the highly complex tasks that are typical for hack week projects. It may hence be advisable to allow for diffusion of the group into adjacent rooms, while providing ample venue for the entire group to congregate for tutorials, breaks and reports.
There should be plenty of power outlets for participants to plug in their computers.
The space should have ready access to wifi, for activities that require an internet connection (e.g., collaborative software development), and organizers should arrange in advance to provide access to visitors from other institutions.
Accessibility should be taken into consideration: is the space accessible for participants who use wheel-chairs? Are accessible restrooms available.

@@

Event size (i.e. the number of attendees) and event duration are important to consider during the design and planning phase of a hack week. All hack weeks held so far have been five days, with group sizes between 40-60 participants. The number of attendees resulted from a number of different design considerations. One practical consideration was the size of the available location: especially the project-focused parts of a hack week generally require a reconfigurable space with more seats than attendees, such that spontaneous organization in sub-groups is easy. However, group size is also important in the context of community building, group dynamics and programme decisions. We limited group size to a maximum of 70 participants, because previous anecdotal evidence has shown us that this is about the maximum group size where interactions between all participants is still possible. We wanted to ensure that all participants had an opportunity to at least meet all other participants, in order to facilitate community building and keep the group from fracturing into cliques. Another important consideration is representation along the many axes of diversity we considered. For example, a single undergraduate participant might feel like an outsider; admitting three or four can mitigate alienation effects. Similarly, in order to foster active collaboration among many different subfields, we ensured admitting participants with many different types of expertise, some of which overlapped. Having a set of participants that is too small would interfere both with these goals of representation and of fostering interdisciplinary collaborations and networking, thus we would advise admitting at least $\sim$30 or so participants if these objectives are an important part of the hack week.

@@

Feedback is a crucial component of a hack week. Because of its experimental design, and the differences in group from year to year, it is important to routinely check whether the workshop still matches up with what participants expect and find useful. Additionally, computational fields tend to be fast-paced; a tutorial given one year might be outdated a few years later. It is thus useful to put some detailed thoughts into the survey. Following from discussions the committee has likely had during the initial stage of preparing the announcement and selecting participants, organizers will want to confront their own expectations before the workshop, and design a survey that allows for both measurements of specific outcomes and goals the organizers are interested in, as well as leave enough free space for participants to self-report outcomes that arose spontaneously or were out of scope of the initial design. It is helpful to design the survey early on and allow for some time near the end of the workshop for participants to fill it out there; in our experience, this leads to much higher response rates than if the survey is distributed after the end of the workshop. While some improvements and changes in the survey design from year to year are inevitable, it is advisable to set a number of core questions that do \textit{not} change from year to year. These questions will allow organizers to assess the workshop's function within their community, and assess how changes to the format between one year and another may affect outcomes of interest.
